---
title: "Climate change in Bergen"
format: html
author: "Ingrid Kregnes"
date: today
editor: visual
bibliography: references.bib
csl: plos.csl
echo: false
---

```{r}
#| label: load-packages
library(tidyverse) # for ggplot2, dplyr, readr etc
library(here)      # for help finding files
library(lubridate) # for handling dates & times
library(gt)        # for tables

# default plot theme
theme_set(theme_bw())

# default number of digit
options(digits = 2)
```

# Introduction

Climate change is expected to have large impacts on both marine [@perry2005] and terrestrial ecosystems in Norway [@topper2018].

# Methods

## Data Analysis

Climate data were downloaded from https://seklima.met.no/

Data were analysed in R version `r paste(R.version$major, R.version$minor, sep = ".")` [@R], using tidyverse for data manipulation and plots [@tidyverse].

# Results

```{r}
#| label: load-data
bergen <- read_delim(
    file = here("data/bergen_weather.csv"), 
    delim = ";", # ; are column delimiters
    locale = locale(decimal_mark = ","), # , separate the decimal from the rest of the number
    col_types = cols(`Tid(norsk normaltid)` = col_date(format = "%m.%Y")),
    na = "-" # missing values denoted by "-"
  ) |> 
  # rename awkward columns
  rename(date = `Tid(norsk normaltid)`, 
         temperature = `Homogenisert middeltemperatur (mnd)`,
         precipitation = `Nedbør (mnd)`) |> 
  # filter out metadata on last line
  filter(!stringr::str_detect(Navn, "Data er gyldig"))

```

```{r}
#| label: model

bergen_annual <- bergen |> 
  mutate(year = year(date)) |> 
  # remove 2022 - only two months of data
  filter(year < 2022) |> 
  group_by(year) |> 
  summarise(annual_mean_temp = mean(temperature))

mod <- lm(annual_mean_temp ~ year, data = bergen_annual)

```

@tbl-monthly-means shows the mean monthly temperatures for Bergen for the period `r min(year(bergen$date))`--`r max(year(bergen$date))`. Mean annual temperatures in Bergen have been increasing over the study period at a rate of `r coef(mod)[2] * 10`°C per decade.

```{r}
#| label: tbl-monthly-means
#| tbl-cap: Mean monthly temperatures in Bergen

bergen |> 
  mutate(month = month(date, label = TRUE)) |> 
  group_by(month) |> 
  summarise(mean_temperature = mean(temperature, na.rm = TRUE)) |> 
  gt() |> 
  fmt_number(columns = mean_temperature, decimals = 1) |> 
  cols_label(month = "Month", mean_temperature = "Temperature °C")
  
```

The maximum annual mean temperature was recorded in `r max(bergen_annual$temperature)`\`

```{r}
#| label: fig-annual-temp
#| fig-cap: Mean annual temperature for Bergen
#| fig-width: 5.2
#| message: false

ggplot(bergen_annual, aes(x = year, y = annual_mean_temp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_line() +
  labs(x = "Year CE", y = "Mean annual temperature °C")
```

```{r}
bergen_annual_precipitation <- bergen |> 
   mutate(year = year(date)) |> 
  # remove 2022 - only two months of data
  filter(year < 2022) |> 
  group_by(year) |>
  summarise(annual_precipitation = sum(precipitation, na.rm = TRUE)) |> 
  filter(annual_precipitation > 0)
  
```

```{r}
ggplot(bergen_annual_precipitation, aes(x = year, y = annual_precipitation)) + geom_point() + labs(x = "Year", y = "Annual precipitation")
```

```{r}
#| label: tbl-monthly-precip

bergen |>
  mutate(month = month(date, label = TRUE)) |>
  group_by(month) |> 
  summarise(mean_precip = mean(precipitation, na.rm = TRUE), sd = sd(precipitation, na.rm = TRUE)) |>
  gt() |>
  cols_label(month = "Month", mean_precip = "Precipiation", sd = "standard deviation")
  
```
